---
title: "Lab6"
output: html_document
---

Topics: Spatial data 
-Raster analysis
-Kriging

Q: Why do raster analysis in R?
-Allows for reproducibility 

geom_sf is best for points and lines
geom_raster() is best for raster

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Attach packages:
```{r}
library(tidyverse)
library(here)
library(janitor)
library(raster)
library(sf)
library(tmap)
library(tmaptools)
library(gstat) # variogram

```

## Grand Canyon GeoTIFF - Data Processing
Load data (raster)
```{r}
gc_dem <- raster(here("data","gc_dem.tif"))

plot(gc_dem)
```

Update coordinate system
```{r}
#Check CRS
gc_dem@crs #WGS84 UTM

#Check the extent (bounding box)
gc_dem@extent #Shows units in m

#Create a WGS84 with latlong metrics
wgs84 <- "+proj=longlat +datum=WGS84 +no_defs" 

gc_dem_reproj <- projectRaster(gc_dem, crs = wgs84, method = "bilinear")

gc_dem_reproj@extent #Now units are in degrees
```

Crop original spatial data
```{r}
#Crop raster to a smaller area (bounding box)
bounds <- as(extent(-112.4, -112.0, 36.1, 36.3), 'SpatialPolygons')

#Make the projection for "bounds" the same as for "gc_reproj":
crs(bounds) <- crs(gc_dem_reproj)

#Crop original data to the new bounding box
gc_dem_crop <- crop(gc_dem_reproj, bounds)
plot(gc_dem_crop)
  
```

Resample to create lower res (larger cell) rasters 
```{r}
gc_dem_agg <- raster:: aggregate(gc_dem_crop, fact = 10) #Higher fact = lower resolution
```
-Use the aggregate() function
-See ?aggregrate (default is mean, `fact` is number of cells in each direction; can set 2 if wanting x/y to differ for aggregation)


Plot in ggplot
```{r}
#Convert raster data to a dataframe (ggplot requires dataframes, not a raster)
gc_df <- as.data.frame(gc_dem_agg, xy = TRUE) #xy = TRUE keeps the lat long coordinates

#Plot in ggplot 
ggplot(data = gc_df, aes(x=x, y=y)) +
  geom_raster(aes(fill = gc_dem)) +
  coord_quickmap() +
  theme_minimal() +
  scale_fill_gradientn(colors = c( #Can also change break points using this function
    "darkblue",
    "blue",
    "yellow",
    "white"
  ))
```

Select cells that match given criteria
```{r}
#Ex: A certain species will only grow between 1000 and 1500 ft elevation. Create a subset of gc_dem_crop that includes that habitat.

#Create a new layer to work on
gc_hab <- gc_dem_crop

#Set any cells outside of (1000 - 1500) to NA
# Uses base R syntax
gc_hab[gc_hab > 1500 | gc_hab < 1000] <- NA 
plot(gc_hab)
```

Make interactive with tmap
```{r}
#Set tmap mode to view
tmap_mode("view") 

tm_shape(gc_hab) +
  tm_raster(legend.show = FALSE, palette = "plasma")

# In console, run tmaptools::palette_explorer() to view a Shiny app with other palettes!
```

## Kriging: Kansas precipitation  

1. Data exploring

Explore the KS counties shapefile 
```{r}
#Read in data
#Use read_sf to create sticky geometries
ks_counties <- read_sf(here("data",
                            "ks_counties",
                            "ks_counties_shapefile.shp"))


#Look at base plot
plot(ks_counties)

#Check CRS:
st_crs(ks_counties) #None... we need to set one

#Set to EPSG 4326 (WGS84 datum):
st_crs(ks_counties) <- 4326

#Now check again:
st_crs(ks_counties)

#And replot:
plot(ks_counties)

#Plot in ggplot
ggplot(data = ks_counties) +
  geom_sf()
```

Explore rainfall data
```{r}
#Read in data
ks_rain <- read_csv(here("data","ks_rain.csv")) %>% 
  clean_names()

#Convert ks_rain data to be recognized as spatial points

ks_sf <- st_as_sf(ks_rain, coords = c("lon", "lat"), #Convert lat/lon columns to sticky geometries 
                 crs = 4326) 


#View(ks_sf) (rainfall = amt)
plot(ks_sf)

#Plot in ggplot
ggplot() +
  geom_sf(data = ks_counties) +
  geom_sf(data = ks_sf, aes(color = amt,
              size = amt),
          show.legend = FALSE) +
  coord_sf() +
  scale_color_gradient(low = "yellow", high = "blue") +
  theme_void()
```

2. Predict rainfall all over Kansas based on our rainfall data

Make a spatial pixel grid to make predictions over
```{r}
#Convert rain sf data to spatial
ks_sp <- as_Spatial(ks_sf)
class(ks_sp) #Make sure class is set to "sp"

#Create lat/long sequences
# bbox(ks_sp) gives you the min and max to help create these sequences
lat <- seq(37, 40, length.out = 200) #Stick with 200 or less otherwise kriging will take forever
long <- seq(-95, -102, length.out = 200)

#Put lat/long sequences into a spatial grid to make predictions over
grid <- expand.grid(lon = long, lat = lat) #Combine lat/long into a grid
grid_sf <- st_as_sf(grid, coords = c("lon", "lat"), crs = 4326) #Convert to sf 

#Convert grid_sf to spatial data
grid_sp <- as_Spatial(grid_sf)

# Check out your amazing kriging-ready grid:
plot(grid_sp)
```


Make a variogram from the rainfall data: 
```{r}
#Create the variogram:
ks_vgm <- variogram(amt ~ 1, data = ks_sp)

#Plot the variogram
plot(ks_vgm)

#Make an estimate for the variogram parameters:
## Nugget = 0.2
## Sill = 0.8
## Range = 200
```

Fit a model to the variogram:
```{r}
#Fit a spherical model to the variogram using reasonable estimates for nugget, sill and range 
ks_vgm_fit <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, model = "Sph", range = 200)) 

plot(ks_vgm, ks_vgm_fit) 

#Can also look at other models (Gaussian, Exponential) 
ks_vgm_gau <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, model = "Gau", range = 200))

plot(ks_vgm, ks_vgm_gau)

#Test for model fit by checking the sum of squares of residuals
attr(ks_vgm_fit, 'SSErr') # 0.00214 
attr(ks_vgm_gau, 'SSErr') # 0.00745

#We'll stick with the Spherical model: 
ks_vgm_fit # Nugget = 0.102, sill = 0.954, range = 235
```


Kriging 
```{r}
#Kriging of grid_sp based on the ks_vgm_fit model 
ks_krige <- krige(amt ~ 1, ks_sp, grid_sp, model = ks_vgm_fit)

#Plot
spplot(ks_krige, "var1.pred")

#Create a new dataframe for the output
ks_df <- data.frame(ks_krige@data["var1.pred"],
                    ks_krige@data["var1.var"],
                    ks_krige@coords) %>% 
  rename(longitude = coords.x1,
         latitude = coords.x2)

#Convert this df to sf object:
rain_sf <- st_as_sf(ks_df, coords = c("longitude","latitude"), crs = 4326)

```

Crop to Kansas
```{r}
#Load Kansas outline
ks <- read_sf(dsn = here("data","states"), 
              layer = "cb_2017_us_state_20m") %>% 
  dplyr::select(NAME) %>% 
  filter(NAME == "Kansas") %>% 
  st_transform(crs = 4326)

plot(ks)

#Find intersection of Kansas outline and rain_sf 
rain_sf_ks <- st_intersection(rain_sf, ks)

#Plot in ggplot
ggplot(rain_sf_ks) +
  geom_sf(aes(color = var1.pred)) +
  theme_minimal()

#Next step: Convert to raster and plot as geom_raster()
```
